---
title: "Final Project"
author: "Jordan Denish and Malcolm Mashig"
date: "5/7/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Generate a publishable Rmarkdown document with the following sections:

# Introduction

For this final project, we want to create an interactive data visualization for a Capstone group project that we worked on throughout the semester. We have developed a Shiny Application on R using the information from class lectures and teaching ourselves how Shiny works online. 

## Question and background information on the data and why you are asking this question(s).  References to previous research/evidence generally would be nice to include.

For our capstone project, we wanted to analyze Major League Baseball, hoping to use historical data and various statistics in order to predict player value and make conclusions about the sport. We narrowed our focus to be more specific, ultimately choosing to analyze and project future performance of starting pitchers. Our inspiration for picking this particular position came from the fact that starting pitchers have become increasingly important and prominent among playoff contending teams. The Washington Nationals, the winner of the 2019 World Series, for example, were anchored by three elite, highly-paid starting pitchers in Max Scherzer, Stephen Strasburg, and Patrick Corbin. 
Furthermore, the size and length of the contracts that teams are giving to starting pitchers are tremendously large. This past winter, during baseball’s free agency, over 1 billion dollar was spent on starting pitchers, the most free agency dollars spent on starting pitchers in baseball history. Free agent Gerrit Cole signed a nine-year, 324 million dollar contract with the New York Yankees, which is the largest deal ever signed by a pitcher. Gerrit Cole is currently 29 years old, which is considered to be one of the prime years for a starting pitcher, but he will be 38 years old at the end of his contract, when most pitchers are already in decline and considering retirement. Only the big market baseball teams like New York, Boston, and Philadelphia have the spending capabilities to afford these massive free agent signings, while the smaller market teams like Tampa Bay and Oakland must find cheap alternatives in order to be able to compete. This relatively recent trend in the importance that the sport is placing on finding valuable starting pitching led us to make it the concentration of our capstone project. 



## Exploratory Data Analysis – Initial summary statistics and graphs with an emphasis on variables you believe to be important for your analysis.

```{r }

```



## Methods – Techniques you are using to address your question and the results of those methods.

```{r }

```


## Conclusions – What can you say about the results of the methods section as it relates to your question.

```{r }

```


## Future work – What additional analysis is needed or what limited your analysis on this project.

```{r }

```
