---
title: "Final Project"
author: "Jordan Denish and Malcolm Mashig"
date: "5/7/2020"
output: 
  html_document:
  theme: cerulean  
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

DS 4001 Final Project: Data Visualization for our Capstone Project

# Introduction

For this final project, we want to create an interactive data visualization for a Capstone group project that we worked on throughout the semester. We have developed a Shiny Application on R using the information from class lectures and by teaching ourselves how Shiny works online. We are very excited about this application because it is much easier for people to use our app than read our 20 page report.

### Question and Background Information

For our capstone project, we wanted to analyze Major League Baseball, hoping to use historical data and various statistics in order to predict player value and make conclusions about the sport. We narrowed our focus to be more specific, ultimately choosing to analyze and project future performance of starting pitchers. Our inspiration for picking this particular position came from the fact that starting pitchers have become increasingly important and prominent among playoff contending teams. The Washington Nationals, the winner of the 2019 World Series were anchored by three elite, highly-paid starting pitchers in Max Scherzer, Stephen Strasburg, and Patrick Corbin. 

Furthermore, the size and length of the contracts that teams are giving to starting pitchers are tremendously large. This past winter, during baseball’s free agency, over 1 billion dollars was spent on starting pitchers, the most free agency dollars spent on starting pitchers in baseball history. Free agent Gerrit Cole signed a nine-year, 324 million dollar contract with the New York Yankees, which is the largest deal ever signed by a pitcher. Gerrit Cole is currently 29 years old, which is considered to be one of the prime years for a starting pitcher, but he will be 38 years old at the end of his contract, when most pitchers are already in decline and considering retirement. Only the big market baseball teams like New York, Los Angeles, and Philadelphia have the spending capabilities to afford these massive free agent signings, while the smaller market teams like Tampa Bay and Oakland must find cheap alternatives in order to be able to compete. This relatively recent trend in the importance that the sport is placing on finding valuable starting pitching led us to make it the concentration of our capstone project. 

After refining our ideas, we gathered data relevant to our project. We took the approach of collecting a large amount of data, hoping this would help us further revise and pinpoint a particular question of interest. A website called FanGraphs was very helpful in this data collection process, containing seemingly every imaginable statistic for every player, separated by season. The site also provided the ability for us to filter the data based on position and various other criteria and easily export the data as a CSV formatted file. We harvested all available starting pitcher data, with two significant constraints. First, we limited how far back our data went to the year 2002. Knowing that only somewhat recent data would be useful for drawing conclusions about the future, coupled with the fact that several advanced baseball metrics, such as xFIP and BABIP, have only been tracked since 2002, it made sense to cut the data off at this year. Our other constraint was that we limited the data to the seasons where the pitcher threw at least 80 innings. Though this admittedly is a somewhat arbitrary cutoff, our experimentation found that this excluded artificially short seasons, usually due to injury, while still providing a large amount of data to analyze. Further, after substantial back and forth consideration about whether to consider injuries, we decided not to pursue this analysis, as the extreme amount of randomness in injuries would likely prevent us from coming away with any meaningful results. Thus, our complete data set included all relevant performance statistics for each season with at least 80 innings pitched for every starting pitcher from 2002-2019. 

### Exploratory Data Analysis

As you can see by the change in fastball velocity and xFIP since 2002, the percentiles have kept a consistent distance. This shows us that standardizing variables like fastball velocity and xFIP will help to mitigate the influence of changes in the game of baseball over time on our analysis. We standardized nearly all of our variables to account for this variation.

```{r, echo=FALSE, warning=FALSE, message=FALSE}
source("Exploration.R")
```

```{r}
plot1
plot3
```

The plots below show the characteristics of the fastball velocity and xFIP variables. We know from them that fastball velocity generally declines as a pitcher ages, which is expected, and that the change in xFIP for players at a certain age has a generally normal distribution. This was the case for players at the age of 28 and at most other ages

```{r}
fangraphs_clean %>% 
      group_by(Age) %>% 
      summarise(avg_fastball_velocity = mean(FBv, na.rm = TRUE)) %>% 
      filter(between(Age, 20, 38)) %>% 
      ggplot() +
      geom_line(aes(Age, avg_fastball_velocity), color = "blue") +
      labs(title = "Fastball Velocity steadily decreases as players age")
fangraphs_clean %>% 
      mutate(change_in_xFIP = lead(xFIP) - xFIP) %>% 
      filter(Age == 28) %>% 
      ggplot() +
      geom_density(aes(change_in_xFIP), fill = "blue") +
      labs(title = "Change in xFIP from age 28 to 29 is normal")
```

### xFIP

One of the most important decisions we had to make as a group was choosing which metric to evaluate pitchers on, as there are many such statistics that each have strengths and weaknesses. We wanted to choose a metric that was descriptive, in that it encapsulates valuable information, and predictive, in that it is relatively stable year-to-year. The traditional metric for measuring pitcher success is Earned Run Average (ERA), an easy to understand statistic which simply divides the number of earned runs a pitcher allows by the number of innings he pitches, and then multiplies this average by nine to mimic the length of a baseball game. ERA makes sense as a critical performance measure, as a pitcher’s primary goal for his team is to prevent runs, and a pitcher’s ERA will show how effective he was. However, ERA has many issues that have caused most baseball analysts to move beyond it. 

For one, the fielding of the team around a pitcher is not included in the metric, which certainly has an impact on the number of runs that he gives up. A pitcher with a better defense is more likely to have a lower ERA than a pitcher with a worse defense. Another negative aspect of ERA is that it includes no information about the ballpark that the pitcher plays in. For instance, Coors Field of the Colorado Rockies is notoriously easy to hit home runs in, as balls fly farther in the high elevation of Denver. On the other end of the spectrum, Petco Park of the San Diego Padres has deep fences and a spacious outfield which are hostile to hitters. Since the environment that pitchers play in varies so much from pitcher to pitcher, a metric like ERA has major flaws that make it difficult to encapsulate a pitcher’s true value. 
Due to these flaws, we eventually decided to use a statistic named 
Expected Fielding Independent Pitching (xFIP), which is an advanced baseball statistic that is an estimate for a pitcher’s earned run average (ERA). A pitcher’s main objective is to prevent runs and xFIP is considered to be a better predictor of future performance than ERA, which can mask the actual 
skill of a pitcher. xFIP is a common estimate for a pitcher’s ERA that only factors in the batter outcomes that a pitcher can control (strikeouts, walks, and hit by pitches) while also normalizing for home runs allowed by using the league average home run/fly ball rate. xFIP is calculated using the formula xFIP = ((13*(Fly balls * lgHR/FB%))+(3*(BB+HBP))-(2*K))/IP + constant. The constant is generally around 3.10, which helps adjust the xFIP value to an ERA scale.

### Research Question

Once we were satisfied and confident in using xFIP as our performance metric, we were able to clearly state our research question: Can we accurately predict an MLB starting pitcher’s future xFIP over the next few seasons using various performance statistics from prior seasons in order to estimate relative future value of the pitcher? Simply put, we wanted to create a model that could take any pitcher, along with the statistics of all of their previous seasons, as an input and make projections for the xFIP of their next few seasons. From our main research question, two sub-questions naturally emerged as well. In particular, we wanted to look at how pitchers “aged”, or how their performance was affected as they got older, as well as whether all prior seasons were equally predictive or if more recent seasons were more important predictors of future xFIP. Our data set provided us with all of the relevant and necessary information to model xFIP over time, as well as answer our sub-questions. Further, viewing our data as a sample of the population of all starting pitcher seasons with at least 80 innings pitched, both past and future, we will be able to generalize our results to future seasons.


<<<<<<< HEAD

### Methods
=======
### Methods – Techniques you are using to address your question and the results of those methods.
>>>>>>> c2e0405f4d90df6eb3664efbd350bd0933a1d061

In order to develop predictions for xFIP, we used a process named lagged linear regression. The theory of our lagged linear regression xFIP model is derived from the idea that in order to generate predictions for the next three years, we would want to produce incremental predictions. If we are looking at a pitcher's performance in 2019, we will first predict xFIP in 2020 and then use that prediction to generate our second prediction in 2021. Finally, we will use the prediction in 2021 to predict the pitcher’s xFIP value in 2022. 

Rather than using this same model to predict multiple years in advance, we used various “submodels” to predict each of these inputs for the next year, and then used these submodel predictions as inputs into our main model to predict the next year’s xFIP. 

For example, we could use additional predictors such as fastball velocity and fastball percentage (percentage of pitches that are fastballs) in 2019 to predict 2020’s xFIP; then, with our submodels, we could project each of the other unknown predictors that are inputs in our main model in 2020. We would then be able to plug these predicted values into our main model to predict 2021 xFIP and continue the process into the future. This is relevant because there were a large number of explanatory variables that could have potentially been used in our model, such as age, years of experience, or pitcher type, and would be known as predictions were made through the future. We can assume age increases by one each year, but we needed submodels that could estimate values for the explanatory variables that change from year to year and depend on additional inputs, such as fastball velocity. 

With this modeling process framework, we were able to predict standardized xFIP for a pitcher multiple years in the future. Once again, we predicted one year in the future, and then used the predicted xFIP value as input for the lagxFIP variable in order to predict the xFIP two years in the future. Each new year out that we predicted, we increased age by one and used the predictions for the last year (fastball velocity, fastball percentage, and xFIP) as input for the next round of submodel and model predictions. We used the model to generate xFIP predictions for the next three seasons, 2020, 2021, and 2022. Going three seasons into the future paints a meaningful picture of the pitcher’s projected performance, and predicting any further resulted in increasingly reduced accuracy and higher variance.

### Model Building

We started with a “submodel” that predicts standardized fastball velocity based on the age of the pitcher and their fastball velocity in the season prior. The model has the following form:

Submodel 1:
FBv = B0 + B1(lagFBv) + B2(age) + B3(young) + B4(prime) + B5(age*young) + B6(age*prime) 
FBv = 0.318 + 0.935(lagFBv) + -0.016(age) + 0.508(young) + 0.278(prime) + -0.020(age*young) + -0.009(age*prime) 

FBv represents the standardized fastball velocity that we want to predict (for the upcoming season). LagFBv represents standardized fastball velocity for the year prior. Age represents the age of the pitcher for the year of the prediction. Since age increases by one for each pitcher each year, we always know age in the year of the prediction. Young represents a dummy variable that is 1 when the player is younger than 25 years old in the year of prediction and 0 otherwise. Prime represents a second dummy variable that is 1 when the player is between the ages of 25 and 31 and 0 otherwise. If both young and prime are 0, then the pitcher is above 31 and classified as old.
Next, we moved onto a “submodel” that predicts standardized fastball percentage based on the age of the pitcher and their fastball percentage in the season prior. The model has the following form:

Submodel 2:
FBp = B0 + B1(lagFBp) + B2(age) + B3(young) + B4(prime) + B5(age*young) + B6(age*prime) 
FBp = -0.094+0.866(lagFBp) +0.001(age) -0.079(young) +0.249(prime) + 0.006(age*young) - 0.007(age*prime)

FBp represents the standardized fastball percentage that we want to predict. LagFBp represents standardized fastball percentage for the year prior. Age, young, and prime have the same meaning as in the fastball velocity submodel.
With the ability to make predictions for fastball velocity and fastball percentage for pitchers in a future year based on the latest available year, we have all the predictors necessary to make predictions for xFIP. We can now create three main xFIP models with the following forms that predict standardized xFIP based on 1,2, or 3 previous years of data.

Model 1:
xFIP= B0 + B1(lagxFIP) + B2(age) + B3(predictedFBv) + B4(predictedFBp) 
xFIP= -0.388 + 0.639(lagxFIP) + 0.014(age) + -0.115(predictedFBv) + -0.0025(predictedFBp) 

Model 2:
xFIP= B0 + B1(lagxFIP) + B2(age) + B3(predictedFBv) + B4(predictedFBp) + B5(lagxFIP2)  
xFIP= -0.324 + -0.014(lagxFIP) + 0.014(age) + -0.102(predictedFBv) + -0.014(predictedFBp) + 0.484(lagxFIP2)  

Model 3:
xFIP= B0 + B1(lagxFIP) + B2(age) + B3(predictedFBv) + B4(predictedFBp) + B5(lagxFIP2) + B6(lagxFIP3) 
xFIP= -0.267 + 0.465(lagxFIP) + 0.012(age) + -0.109(predictedFBv) + -0.007(predictedFBp) + 0.264(lagxFIP2) + 0.036(lagxFIP3) 

### Analysis

```{r }
TestError <- data.frame(c(0.3516178,0.4445261,0.4783301
)) 

rownames(TestError) = c("Year1ME", "Year2ME", "Year3ME")

TestError
```

The table above displays our average residuals from using our training model to predict test data xFIP values. The above errors represent the average distance that the predictions are away from the actual xFIP values for each of the three seasons in the future. This means that on average, we would expect our test predictions for years 2020, 2021, and 2022 to be off by about 0.35, 0.44, and 0.48 from the actual xFIP values respectively. We are pleased that these values are not much higher than our training absolute error, which means that we would expect our model to perform almost as well when predicting future xFIP values than the existing xFIP values. 


## Conclusions

Through analyzing our model’s predictions, we have been able to draw a number of conclusions about starting pitchers in baseball. An important inference we were able to draw is that the typical baseball pitcher’s career follows an arc in which their performance steadily improves in their early years in the league, plateaus in their prime, around the 27 to 31 age range, and quickly declines in their later years, often even “falling off a cliff”. Our model also provided a good, real example of the fact that while outliers certainly do exist, regression to the mean is much more prominent. Essentially, we should not expect pitchers to remain excellent for extended periods of time because it is really difficult to do. However, this also means that pitchers who can consistently maintain excellence should be perceived as extremely valuable. 

In our analysis, we also discovered that one’s performance from the past year is overwhelmingly more important than prior performance and any other potential predictor. This makes sense for most players, but it somewhat debunks the idea that a player just had a “down year” and will return to their previous form, as their down year seems to actually be more indicative of subsequent performance. Looking at our predictions for Gerrit Cole, who signed a nine-year $324 million contract this past offseason and served as part of the inspiration for our project, provides an interesting example of our model in action. Our model predicts Cole to continue his dominance and be the best pitcher in baseball for each of the next three years, even in 2022 when he will turn 32. This means that our model projects Cole to have the league’s lowest xFIP value during each of the next three seasons. Though it is almost inevitable that Cole will fall off before the end of his contract, when he will be 38, even three years of Cy Young caliber performance could justify such a massive deal, especially if it pushes the Yankees over the hump to win a World Series.

Another conclusion we were able to draw is that fastball velocity shares a very strong relationship with performance. Fastball velocity was the most important predictor of performance behind the xFIP values included in the model, and our submodel defined a strong negative linear effect of age on the change in fastball velocity year to year. It is no surprise that as a player’s physical skills wane, his average fastball velocity decreases. It is also clear in our analysis that harder throwing pitchers generally have better xFIP values. That said, it is interesting to see that as a pitcher ages, his velocity decreases, and xFIP tends to decline alongside a declining fastball velocity. It is very possible that performance is influenced by a chain of cause and effect in this way. Even if that is not the case, our analysis shows that it is very important for teams to evaluate how the speed of a pitcher’s fastball is changing, because it could very well have some indication for how they will perform through the future.

Additionally, we also learned that predicting baseball is a very complicated concept. While we are able to estimate xFIP values with reasonable accuracy, there would need to be many more model adjustments to allow a baseball general manager to make decisions about player personnel. Teams have to assess the health of the athletes, the potential variability of their performance, and their importance to the franchise as a teammate and role model. If a pitcher is expected to have an extremely low xFIP but only stays healthy for half of the season, how should that pitcher compare to an average pitcher who has stayed healthy his entire career? General managers have to be able to assess the talent that is available in the league at each position, and the replacement level value of pitchers in each team’s farm system or in free agency.


### Future work

If we were to build upon our analysis further, there would be certain directions in which we would want to move forward. The baseball world is currently on hold due to the Coronavirus pandemic, but once baseball returns, it would be very valuable to evaluate our predictions for 2020 against the real season. We would have the opportunity to adjust our model appropriately. The game of baseball is constantly changing, and we expect our model to continuously evolve so that it can be as accurate as possible.

Our main inspiration for this project was the increasing size and length of contracts given to starting pitchers. Initially, we hoped to build predictive models of performance in order to infer a player’s fair contract value. Players are offered contracts that guarantee a certain amount of money over many years, and it is important for teams to evaluate the expected performance of a pitcher throughout their entire contract. While a stellar pitcher like Gerrit Cole might be worth 36 million dollars (or even more) this year, his level of performance may steeply decline throughout his nine year contract. He will likely be worth well below 36 million dollars when he is 38 years old. It is important to weigh a player’s expected future performance against the contract they are offered. Ultimately, as a next-step, we would want to evaluate the fairness of contracts based on the predictions our model makes. This would add a lot more nuance to the project because contracts are not just based on player performance. Other aspects such as past injuries, “star” value, and the signing team’s championship hopes are surely large factors. That said, in order to accurately assess contract value, we would have to gather additional data and include these factors in our models.
	
### Our app

After generating our predictions and tinkering with our model, the final step of our project was to find a valuable way to display our findings. We wanted something that was easy to use but also far beyond simply publishing our final dataframe online. Ultimately, we hoped our final deliverable would mimic a baseball card, where the user would be able to view basic information about each player, including our future predictions. Although we collectively had little experience with it, we decided to try to build a Shiny application, as we could develop and run the application in R where the remainder of our code was. Fortunately, it was relatively easy to get our dataframe of predictions to display in the app, as well as menus which allow the user to subset to a specific player or team. After this basic functionality, we began to brainstorm and implement many other features that boosted the usability and value of the app. We were able to scrape biographical information and a picture of every player from FanGraphs and mlb.com, and easily put these objects onto the player pages of the app. Additionally, we displayed graphs that showed past and predicted xFIP values to the player and team pages. One of the most interesting features that we added to the app was an xFIP calculator, where the user can plug in values and our model will give predictions for future xFIP based on the inputted variables. Ultimately, developing our Shiny app was a very smooth process. The seamless integration that it provides with R enabled us to easily upload our dataframes and create interesting graphs without having to build a website from the ground up. We are all extremely pleased with how the app turned out from an aesthetic and usability standpoint, and think that people with limited or advanced baseball knowledge can gain valuable information from using it. 

Our HTML link to the app is https://malcolmmashig.shinyapps.io/Predicting-MLB-Pitcher-Value/